---
title: "Intro Rmarkdown"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r percip}
data("precip")
# summary(precip)
# View(precip)
```

#Outcomes
1.Create HTML report using Rstudio
2.Embed plots in HTML
3.Manipulate named vectors
#Instructions
1.In this homework, we’ll analyze R’s included precip dataset, which contains “Annual Precipitation in US Cities”.\n
2.Answer the following questions, and show the R code you used to answer questions 2-6.\n
3.Upload your submission to Canvas in nicely formatted HTML generated from Rstudio.\n
4.Hint: The following functions will help you- names, sort, head, tail, hist, %in%\n
#Questions
1.Rainfall is typically measured in units of length. Cross reference an external source to find out what units this data set is in. Cite your source with a hyperlink, as described in the web page (Links to an external site.).
[precip data set information](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/precip.html)

2.Verify Sacramento is in the data.
```{r}
# names(precip)[6]
"Sacramento" %in% names(precip)
```

3.Which 5 cities receive the most rain? The least rain?
```{r}
data = sort(precip, decreasing = TRUE)
print(head(names(data)))
print(tail(names(data), n=5))
```

4.How many times larger is the rainfall in the rainiest city compared to the least rainy city?
```{r}
first = data[1]
last = tail(data, n=1)
first[1]/last[1]
```

5.Which city is closest to the mean rainfall? The farthest from the mean rainfall?

```{r}
plot(data)
mean(data)
closest = data[42]
closest
furthest = 67-mean(data)
furthest
```

6.Plot a histogram of precipitation. Interpret this histogram in one or two sentences.
most cities have rainfall between 30-50 inches per year
```{r}
hist(precip)
```

7.The following plot represents this data. What other data are necessary to create this plot?
latitude and longitude of each city
8.What general patterns in annual precipitation does the following plot suggest?
It rains more in the south and the northeast














[link](https://stackoverflow.com/questions/29787850/how-do-i-add-a-url-to-r-markdown)

