---
title: "Intro Rmarkdown"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r percip, echo=FALSE}
data("precip")
```


1.Rainfall is typically measured in units of length. Cross reference an external source to find out what units this data set is in. Cite your source with a hyperlink, as described in the web page (Links to an external site.).

  - The units are in inches: [precip data set information](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/precip.html)

2.Verify Sacramento is in the data.
```{r}
# names(precip)[6]
"Sacramento" %in% names(precip)
```

3.Which 5 cities receive the most rain? The least rain?

  - Mobile, Miani, San Juan, New Orleans and Juneau have the most rain.
  
  - Boise, Albuquerque, El Paso, Reno and Phoenix have the least rain. 
```{r}
data = sort(precip, decreasing = TRUE)
print(head(names(data), n=5))
print(tail(names(data), n=5))
```

4.How many times larger is the rainfall in the rainiest city compared to the least rainy city?

  - Mobile has 9.57 times as much rainfall as Phoenix.
```{r}
first = data[1]
last = tail(data, n=1)
first[1]/last[1]
```

5.Which city is closest to the mean rainfall? The farthest from the mean rainfall?

  - The closest is the value which has smallest distance from the mean. Clevelend is closest to the mean. 
```{r}
which.min(abs(precip - mean(precip)))
```
  

  - The  furthest is the greater of the difference between the mean and the first or  last element in the ordered list. Since the last element is 7 (max) and the first  element is 67  (min) and the mean is 34.89, the 67 is further from the mean than 7 so the last element is the furthest from the mean. Mobile is the furthest from the mean. 

```{r}
mean = mean(precip)
first-mean
mean-last

```

6.Plot a histogram of precipitation. Interpret this histogram in one or two sentences.
Most cities have average annual rainfall between 30-50 inches.
```{r}
hist(precip)
```

7.The following plot represents this data. What other data are necessary to create this plot?

  - The missing data to create this plot is latitude and longitude of each city.

8.What general patterns in annual precipitation does the following plot suggest?

  - It rains more in the south and the northeast. 





