---
title: Predicting the interest rate
date: 8 November 2020
---

Our goal in this assignment is to predict the interest rate of government debt as accurately as possible.


#### Outcomes

- Fit statistical learning models to multivariate data
- Create numeric variables from character vectors
- Compare test set performance of different models


#### Instructions

- Answer the following questions, and show all your R code.
- Upload your submission to Canvas in nicely formatted HTML generated from Rstudio.


### Data

_The [California DebtWatch](https://data.debtwatch.treasurer.ca.gov/Raw-Data/CDA-ALL-Raw/x7jp-pweb) contains the following information:_

> The principal amounts, sale dates, interest rates, terms, purposes, ratings, costs of issuance, financing team participants, issuance documents, and annual reporting (if applicable), among 67 other data points required under California Government Code section 8855, of the various types of debt issued by all state and local government agencies in California.

_Download the data in CSV format and load it into R.
Randomly select 20% of the rows and save them into a test set to use later to evaluate the performance of your model.
Which column(s) represent interest rate?_


```{r}
set.seed(280)
data = read.csv("/Users/austinwilson/coding/r/assignment9/CDA_ALL_Raw.csv")
data2 = read.csv("/Users/austinwilson/coding/r/assignment9/CDA_ALL_Raw.csv")
# randomize data
data = data[sample(nrow(data)),]
# size of the train and test set
train_size = floor(dim(data)[1] * .8)
test_size = dim(data)[1] - train_size
# subset data to get train and test sets  
train = head(data,train_size)
test = tail(data,test_size)
# find interest rate 
interest = grep("Interest", colnames(data), value = TRUE)
interest
```


The columns TIC.Interest.Rate and NIC.Interest.Rate represent the interest rate. I used TIC.Interest.Rate because this represents true interest rate.  

### Calculated Columns

_Define one or more new columns from existing text columns in the data set.
For example, you could add a logical column indicating whether the term "lease" appears in some column.
Why do you think this new column will help you improve the accuracy of your model?_


```{r}
data$lease = grepl("lease", data$Debt.Type, ignore.case = TRUE)
test$lease = grepl("lease", test$Debt.Type, ignore.case = TRUE)
train$lease = grepl("lease", train$Debt.Type, ignore.case = TRUE)
```

- This will help improve the accuracy of you model because it will tell you weather the loan was a lease or not. A lease may have a lower interest rate because the capital is not owned by the debtor. In my case I am only using numerical columns in my model so this column will not be used. 

### Models

_Use the remaining 80% of the data (the training set) to come up with two different models to predict interest rate.
You're welcome to use any external machine learning libraries you like, or you can stick with the `lm` and `rpart` from class.
Note that you can come up with different models by using different subsets of columns.
For example, a model with 3 input columns differs from a model with 60 input columns.
Briefly describe the two models you ended up with._


```{r}
# this is just to print out which columns are numeric 
numeric_columns = c()
for (column in colnames(data)) {
  if (class(data[,column])[1] == "numeric") {
    numeric_columns = c(numeric_columns,column)
  }
}
numeric_columns
numeric_columns_with_most_data = c("TIC.Interest.Rate", "Principal.Amount", "New.Money", "Refunding.Amount", "Issue.Costs.Pct.of.Principal.Amt", "Total.Issuance.Costs", "UW.Total.Discount.Spread")
```


```{r}
# numeric columns only 
test_numeric = test[,numeric_columns]
test_numeric = na.omit(test_numeric)

train_numeric = train[,numeric_columns]
train_numeric = na.omit(train_numeric)

model0 = lm(TIC.Interest.Rate ~ Principal.Amount, data = train_numeric)
model1 = lm(TIC.Interest.Rate ~ ., data = train_numeric)
library("rpart")
model2 = rpart(TIC.Interest.Rate ~ ., data = train_numeric)
model3 = lm(TIC.Interest.Rate ~ Principal.Amount + NIC.Interest.Rate + NIC.Interest.Rate + Issue.Costs.Pct.of.Principal.Amt + Total.Issuance.Costs +  UW.Total.Discount.Spread + Credit.Enhancement.Fee, data = train_numeric)

summary(model0)
summary(model1)
summary(model3)
# model2
```


- The two models I came up with were one with all the numeric columns and one that took the best predictors. I used the p-value to eliminate a couple of columns and improve my models accuracy. I will not list out all the columns for the initial model. I also tested a model that just used principle amount (a good predictor of interest rate) which was much less accurate than my final model. The final model used Principal.Amount, NIC.Interest.Rate, NIC.Interest.Rate, Issue.Costs.Pct.of.Principal.Amt, Total.Issuance.Costs, UW.Total.Discount.Spread, Credit.Enhancement.Fee to predict interest rate. 


### Evaluating Performance

_Evaluate both of your models on the 20% of the data you reserved for the test set by looking at the average absolute difference between the interest rate predicted by the model and the actual interest rate.
Do the models do a reasonable job of predicting interest rate?
Find the rows where the predicted interest rate is farthest from the true interest rate.
Why might the model have done a poor job on these rows?_




```{r}

average_absolute_difference = function(predicted, actual) {
  absolute_difference = abs(predicted- actual)
  mean(absolute_difference)
}

abs_difference = function(predicted, actual) {
  abs(predicted- actual)
}



prediction0 = predict(model0, test_numeric)
add0 = average_absolute_difference(prediction0, test_numeric$TIC.Interest.Rate)
test_numeric$prediction0 = prediction0
test_numeric$difference0 = abs_difference(prediction0, test_numeric$TIC.Interest.Rate)


test_numeric$mother_russia = abs(prediction0 - test_numeric$TIC.Interest.Rate)

prediction1 = predict(model1, test_numeric)
aad1 = average_absolute_difference(prediction1, test_numeric$TIC.Interest.Rate)
test_numeric$prediction1 = prediction1
test_numeric$difference1 = abs_difference(prediction1, test_numeric$TIC.Interest.Rate)



prediction2 = predict(model2, test_numeric)
aad2 = average_absolute_difference(prediction2, test_numeric$TIC.Interest.Rate)
test_numeric$prediction2 = prediction2
test_numeric$difference2 = abs_difference(prediction2, test_numeric$TIC.Interest.Rate)


prediction3 = predict(model3, test_numeric)
aad3 = average_absolute_difference(prediction3, test_numeric$TIC.Interest.Rate)
test_numeric$prediction3 = prediction3
test_numeric$difference3 = abs_difference(prediction3, test_numeric$TIC.Interest.Rate)



add0
aad1
aad2
aad3


test_numeric = test_numeric[order(test_numeric$difference3),]


tail(test_numeric)


# these are the rows that we could not accurately predict (furthest from the actual interest rate)
# 27674
# 35262
# 23857
# 50954
# 33339
# 64188
data[27674,]
data[35262,]
data[23857,]
data[50954,]
```


- These models do a reasonable job of predicting interest rate. The first model was the least acurate because it only depended on principle amount. I then made a model with all of the numeric columns. I looked at the summary of this model and selected only the columns that had a small p-value. This model was the most accurate of them all. 

- For the one that was furthest away (row 2674) this may be innacurate because the source of repayment was private obligor payments. This could potentially cause a higher interest rate because a private authority may be more likely to default on a payment. 

- For the second least accurate row (35262) it may be because this was a lease. It seems that the lease column did come in handy, although this is purely speculation. 

- The next least accurate was for Davis Joint Unified School District! I don't see any reason why this would be inaccurate. 

- The next was also using private obligor payments as the repayment source which I think may have contributed to the inaccuracy. 

