---
title: Covid Cases
date: 11 October 2020
---

#### Outcomes

- Download and load public data sets
- Read and interpret metadata
- Evaluate general claims using data, in the presence of missing data
- Create and interpret interaction plots


#### Instructions

- Answer the following questions, and show all your R code.
- Upload your submission to Canvas in nicely formatted HTML generated from Rstudio.

Download the [COVID-19 Case Surveillance Public Use Data](https://data.cdc.gov/Case-Surveillance/COVID-19-Case-Surveillance-Public-Use-Data/vbim-akqf) from the US Center for Disease Control (CDC).
Click on the "Export" button in the top right corner next to download the actual data set.

_Hint: It might be useful to create a new numeric column for severity based on the values of the `hosp, icu, death` columns.
If you do this, make sure you explain the meaning and rationale behind your choices._


# Questions

## 1 - Reading the Documentation

_a) Describe this data set at a high level.
What does each row represent?_

Each row represents a patient. 

_b) How many rows and columns do you expect to see based on the website? 
Load the data into R.
Is this reflected in the data? 

```{r}
covid_data = read.csv("COVID-19_Case_Surveillance_Public_Use_Data.csv")
dim(covid_data)
row_count = dim(covid_data)[1]
```

- I don't see any information on the website about the data set dimensions. The size of the file is 462 MB so I would expect a large dataset, with thousands of rows and a couple columns. 

- Yes. The data set has 4,481,062 rows and 11 columns. More than I anticipated. 


## 2 - Missing Data
 
_There are few `NA` values in this data, but much of the data is missing, unknown, or blank.
Which columns have the highest proportion of missing data?
Why might these be missing?_

```{r}


# sex missing values
count_missing = function(data) {
  row_count = dim(data)[1]
  for (column in colnames(data)) {
    # "Missing"
    missing = sum(data[,column] == "Missing", na.rm = TRUE)
    # unknown
    unknown = sum(data[,column] == "Unknown", na.rm = TRUE)
    # "N/A"
    na = sum(is.na(data[,column]))
    # blank or ""
    blank = sum(data[,column] == "", na.rm = TRUE)
    total = sum(missing,unknown,na,blank)
    percent_missing = total / row_count
    result = c(column, total, total/row_count)
    print(result)
    # print(column)
    # print(total)
    # print(total/row_count)
  }
  
}

count_missing(covid_data)
# should have used lapply 


```

- icu_yn has the most values "Missing", "Unknown" blank or NA. medcond_yn has the second most missing values. pos_spec_dt is the third.  These might be missing because they are unknown or the data was unavailable. There is also probably a lot of missing data because the situation is hectic. Doctors may not have time to record this data. If a patient is feeling well they may leave unexpectedly. 


## 3 - Age Risk

_The [CDC states](https://www.cdc.gov/coronavirus/2019-ncov/need-extra-precautions/older-adults.html) "Among adults, the risk for severe illness from COVID-19 increases with age, with older adults at highest risk."
Does the data support this claim?_

_Hint 1: It might be useful to think about these COVID illness rates by age after adjusting for the relative sizes of each age group in the country.
You'll need to bring in another data source to do this._


```{r}
# from the notes
covid = read.csv("COVID-19_Case_Surveillance_Public_Use_Data.csv")

severity = rep(0, times = nrow(covid))
severity[covid$hosp_yn == "Yes"] = 1
severity[covid$icu_yn == "Yes"] = 2
severity[covid$death_yn == "Yes"] = 3

covid$severity = severity


# young 
young_adults = covid[(covid$age_group == "20 - 29 Years" | covid$age_group == "30 - 39 Years"),]

# dim(young_adults)
print("Young adults:")
dim(young_adults[young_adults$severity == 0,])[1] / dim(young_adults)[1]
dim(young_adults[young_adults$severity == 1,])[1] / dim(young_adults)[1]
dim(young_adults[young_adults$severity == 2,])[1] / dim(young_adults)[1]
dim(young_adults[young_adults$severity == 3,])[1] / dim(young_adults)[1]



# middle aged adults
middle_adults = covid[(covid$age_group == "40 - 49 Years" | covid$age_group == "50 - 59 Years"),]
# dim(middle_adults)
print("middle adults")
dim(middle_adults[middle_adults$severity == 0,])[1] / dim(middle_adults)[1]
dim(middle_adults[middle_adults$severity == 1,])[1] / dim(middle_adults)[1]
dim(middle_adults[middle_adults$severity == 2,])[1] / dim(middle_adults)[1]
dim(middle_adults[middle_adults$severity == 3,])[1] / dim(middle_adults)[1]

# elderly 
elder_adults = covid[(covid_data$age_group == "60 - 69 Years" | covid_data$age_group == "70 - 79 Years" | covid_data$age_group == "80+ Years"),]
dim(elder_adults)
print("elder adults")
dim(elder_adults[elder_adults$severity == 0,])[1] / dim(elder_adults)[1]
dim(elder_adults[elder_adults$severity == 1,])[1] / dim(elder_adults)[1]
dim(elder_adults[elder_adults$severity == 2,])[1] / dim(elder_adults)[1]
dim(elder_adults[elder_adults$severity == 3,])[1] / dim(elder_adults)[1]




```


This data shows that as age trends upward people in category 0 trends downward and people in category 3 trend upward. The claim that the risk for severe illness increases with age. 



```{r}
covid = covid[covid$age_group != "Unknown",]
severity_table = table(covid$age_group, covid$severity)
plot(severity_table)
```




```{r}
# TODO
# unique(covid_data$age_group)




young_boolean_mask = covid_data$age_group == "0 - 9 Years" | covid_data$age_group ==  "10 - 19 Years" 
young = covid_data[young_boolean_mask,]
young_dead = young[young$death_yn == "Yes",]
dim(young_dead)[1] / dim(young)[1]



mid_boolean_mask = covid_data$age_group == "20 - 29 Years" | covid_data$age_group == "30 - 39 Years" | covid_data$age_group == "30 - 39 Years" | covid_data$age_group == "40 - 49 Years"
mid = covid_data[mid_boolean_mask,]
mid_dead = mid[mid$death_yn == "Yes",]
dim(mid_dead)[1] / dim(mid)[1]




old_boolean_mask = covid_data$age_group == "50 - 59 Years" | covid_data$age_group == "60 - 69 Years" | covid_data$age_group == "70 - 79 Years" | covid_data$age_group == "80+ Years"
old = covid_data[old_boolean_mask,]


old_dead = old[old$death_yn == "Yes",]

dim(old_dead)[1] / dim(old)[1]



```


The three values printed are death rates of young, middle aged and elder people, respectively. We can see that as age increases so does the death rate. Therefore this data supports the claim that 








## 4 - Medical Condition Risk

_The [CDC states](https://www.cdc.gov/coronavirus/2019-ncov/need-extra-precautions/people-with-medical-conditions.html) "People of any age with certain underlying medical conditions are at increased risk for severe illness from COVID-19"
Does the data support this claim?_

```{r}
# TODO
# unique(covid_data$medcond_yn)
boolean_mask_medcond_death = (covid_data$medcond_yn == "Yes" | covid_data$medcond_yn == "No") & (covid_data$death_yn == "Yes" | covid_data$death_yn == "No")
medcond_alive_or_dead = covid_data[boolean_mask_medcond_death,]

dim(medcond_alive_or_dead)




# has medical condition and is alive
has_cond_and_alive = medcond_alive_or_dead[(medcond_alive_or_dead$medcond_yn == "Yes" & medcond_alive_or_dead$death_yn == "No"),]
print("percent of people with medical condition who survived")
dim(has_cond_and_alive)[1] / dim(medcond_alive_or_dead)[1]


# has medical condition and is dead
has_cond_and_dead = medcond_alive_or_dead[(medcond_alive_or_dead$medcond_yn == "Yes" & medcond_alive_or_dead$death_yn == "Yes"),]
print("percent of people with medical condition who died")
dim(has_cond_and_dead)[1] / dim(medcond_alive_or_dead)[1]


# does not have medical condition and is alive
no_cond_and_alive = medcond_alive_or_dead[(medcond_alive_or_dead$medcond_yn == "No" & medcond_alive_or_dead$death_yn == "No"),]
print("percent of people with NO medical condition who survived")
dim(no_cond_and_alive)[1] / dim(medcond_alive_or_dead)[1]

# does not have medical condition and died
no_cond_and_dead = medcond_alive_or_dead[(medcond_alive_or_dead$medcond_yn == "No" & medcond_alive_or_dead$death_yn == "Yes"),]
print("percent of people with NO medical condition who died")
dim(no_cond_and_dead)[1] / dim(medcond_alive_or_dead)[1]

# med_cond_table = table(medcond_alive_or_dead$medcond_yn,medcond_alive_or_dead$death_yn)
covid4 = covid[(covid$medcond_yn == "Yes" | covid$medcond_yn == "No"),]
med_cond_table = table(covid4$medcond_yn,covid4$severity)

plot(med_cond_table, xlab = "medical condition", ylab = "death")

```

I subsetted the data so that I only have entries that contain data on medical condition and weather or not  they died. We  can see  that more people who have a medical condition die. Because the numbers in this data seem a little of I created a table and plotted it just to be sure. From the plot of the table I can see that the people with medical conditions much more often fall into categories 1,2 and 3 than people without a medical condition. This supports the claim people with medical conditions are at higher risk. It is worthwhile to mention that this may be biased. People with medical conditions are more likely to check themselves into the hospital. 




## 5 - Interaction Effects

_In statistics, an __interaction effect__ is when two or more variables have a non additive effect in the value of some response.
In the questions above, we explored the severity of COVID-19 (response) given variables age and medical condition.
Do age and medical condition have an interaction effect in the severity of COVID-19?_

```{r}
covid5 = covid[(covid$medcond_yn == "Yes" | covid$medcond_yn == "No"),]
interaction.plot(covid5$age_group, covid5$medcond_yn,covid5$death_yn == "Yes", xlab = "age", ylab = "mean death", trace.label = "medical condition", col =  c("green", "red"))
```



Yes age and medical condition have an effect on severity of covid-19. We can see that as age increases deaths increase. We can also see that in every age category someone with with a medical condition is more likely to die. 


## 6 - Your Question

_Come up with your own question about COVID-19, and answer it using this data set.
Does sex or race (white/non-white for simplicity) have any effect on the death of people. 

```{r}
mask6 = (covid_data$sex == "Male" | covid_data$sex == "Female") & (covid_data$death_yn == "Yes" | covid_data$death_yn == "No")
covid6 = covid_data[mask6,]
covid_differences_in_sex = table(covid6$sex, covid6$death_yn)
plot(covid_differences_in_sex)
```

We can see from this plot that there is no difference, or very little difference in death rates between women and men. 

```{r}
# race 
covid_data$white = covid_data$Race.and.ethnicity..combined. == "White, Non-Hispanic"
covid_differences_in_race = table(covid_data$white, covid_data$death_yn)
plot(covid_differences_in_race)



unique(covid_data$Race.and.ethnicity..combined.)
```

It appears that covid is not racist, and covid does not affect people different if they are white or not. 























